---
title: "Authentication Tutorial"
author: "Adam Varley"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(NatureCubeR)
```

This document explains how to authenticate with the NatureCube API using `NatureCubeR`.

## Overview

All `NatureCubeR` functions that call the API require an `hdr` object — a list
containing your API key and a configured base URL. The two-step process is:

1. Retrieve your API key with `get_key()`
2. Build the auth object with `auth_headers()`

---

## Prerequisites

Obtain your API key from the NatureCube dashboard at <https://naturecube.io>.

Store your credentials in `~/.Renviron` so they are available to every R
session without hard-coding them in scripts:

```
NATURECUBE_API_KEY=your_api_key_here
NATURECUBE_URL=https://naturecube.io/api/
```

You can open this file quickly with:

```{r edit-renviron, eval=FALSE}
usethis::edit_r_environ()
```

Restart R after saving so the variables are loaded.

---

## Step 1 — Retrieve your API key

`get_key()` reads `NATURECUBE_API_KEY` from the environment. If the variable
is not set it raises an informative error.

```{r get-key, eval=FALSE}
api_key <- get_key()
```

You can also supply the key directly — useful for a one-off test, but avoid
committing hard-coded keys to version control:

```{r get-key-explicit, eval=FALSE}
api_key <- get_key(api_key = "your_api_key_here")
```

---

## Step 2 — Build auth headers

`auth_headers()` combines your key with the base URL and returns the `hdr`
list expected by all other `NatureCubeR` functions.

```{r auth-headers, eval=FALSE}
hdr <- auth_headers(api_key)
```

The base URL is read from `NATURECUBE_URL`, falling back to
`https://naturecube.io/api/` if the variable is not set. You can also pass it
explicitly:

```{r auth-headers-explicit, eval=FALSE}
hdr <- auth_headers(api_key, NATURECUBE_URL = "https://naturecube.io/api/")
```

---

## Step 3 — Verify the connection

`get_project()` is a lightweight call that returns basic project metadata.
A successful response confirms that your key and URL are correct.

```{r verify, eval=FALSE}
project <- get_project(hdr = hdr)
print(project)
```

---

## Development / SIT environment

Use `auth_headers_dev()` when working against the staging environment. It
reads `NATURECUBE_SIT_URL`, defaulting to
`https://sit.api.naturecube.io/api/`.

Add to `~/.Renviron` if you need a non-default SIT URL:

```
NATURECUBE_SIT_URL=https://your-sit-url/api/
```

```{r auth-dev, eval=FALSE}
hdr_dev <- auth_headers_dev(api_key)
project_dev <- get_project(hdr = hdr_dev)
```

---

## Full tutorial script

See `tutorials/auth_tutorial.R` for a complete, runnable example.
