---
title: "Observation Upload From CSV"
author: "Adam Varley"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(okalaR)
```

This document describes the easiest route to upload observations to the
`uploadObservations` endpoint.

## What the workflow does

- Pulls project schema once using your API key
- Selects a system and procedure
- Resolves `item_uuid` from `item_name` when needed
- Groups rows into observations using:
  - `observation_id` when present
  - otherwise `(longitude, latitude, recorded_at)`
- Uploads all built observations

## Required CSV columns

Minimum required columns:

- `longitude`
- `latitude`
- `recorded_at`

Recommended columns:

- `item_uuid` (optional if `item_name` is present and schema has matching names)
- `item_name`
- `data` (text values)
- `numbers` (numeric fallback)
- `observation_id` (optional, improves grouping)

> **Note:** only simple point observations are currently supported.
> Nested sub-observations, polygons, and lines are not yet supported.

---

## Step 1 — Authenticate

See `tutorials/auth_tutorial.Rmd` for full details. In short:

```{r auth, eval=FALSE}
api_key <- get_key()
hdr     <- auth_headers(api_key)
```

---

## Step 2 — Fetch project systems and select a procedure

`get_project_systems()` returns the full schema for your project.
`list_systems()` prints a readable summary of available systems and procedures.
`get_procedure()` extracts the specific procedure you want to upload against.

```{r project-systems, eval=FALSE}
project_systems <- get_project_systems(hdr)

# Print available systems and procedures.
list_systems(project_systems)

# Select the procedure that matches your CSV data.
procedure <- get_procedure(
  project_systems,
  system_name    = "Plante Ivindo",
  procedure_name = "Arbre"
)
```

---

## Step 3 — Validate the CSV

Run this before uploading to catch missing columns or formatting issues early.

```{r validate, eval=FALSE}
csv_path <- "tutorials/data/example_observation_data.csv"

validate_csv_against_procedure(
  procedure = procedure,
  csv_path  = csv_path
)
```

---

## Step 4 — Dry run then upload

Always dry-run first. Check that `unresolved_rows` is zero before uploading.

```{r dry-run, eval=FALSE}
result <- upload_observations_from_csv(
  hdr                = hdr,
  csv_path           = csv_path,
  procedure          = procedure,
  dry_run            = TRUE,
  recorded_at_format = "%d/%m/%Y %H:%M"
)

nrow(result$unresolved_rows)
```

If unresolved rows are zero, set `dry_run = FALSE` to upload:

```{r upload, eval=FALSE}
upload_observations_from_csv(
  hdr                = hdr,
  csv_path           = csv_path,
  procedure          = procedure,
  dry_run            = FALSE,
  recorded_at_format = "%d/%m/%Y %H:%M"
)
```

---

## Full tutorial script

See `tutorials/upload_observations_from_csv.R` for a complete, runnable example.
