---
title: "Observation Upload From CSV"
author: "Adam Varley"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(NatureCubeR)
```

This document describes the easiest route to upload observations to the
`uploadObservations` endpoint.

## What the workflow does

- Pulls project schema once using your API key
- Selects a system and procedure
- Validates a wide-format table (one row per feature, procedure item names
  as column headers) against it, resolving item UUIDs from item names
- Uploads only the rows that pass validation

## Required CSV columns

Minimum required columns:

- `longitude`
- `latitude`
- `timestamp` (any format `lubridate::as_datetime()` can parse, e.g. `2026-04-15T08:39:00Z`)
- One column per procedure item, named to match that item's name in the
  schema (e.g. a `DBH` item needs a `DBH` column)

If the procedure has a `label` data_type item, add taxonomic rank columns -
`species`, `genus`, `family`, `order`, `class`, `phylum`, `kingdom` - looked
up by those fixed names, not by the item's own name. Not every organism is
identified all the way to species: the most specific non-blank rank is used
(e.g. one only identified to genus is still uploaded, using that genus).
Each unique resulting value is checked once against the wider IUCN species
database.

Every observation is validated as a whole: a bad timestamp, non-decimal
lon/lat, missing media file, unmatched label, or wrong-typed value rejects
that entire observation rather than uploading it partially.

> **Note:** only simple point observations are currently supported.
> Nested sub-observations, polygons, and lines are not yet supported.

---

## Step 1 — Authenticate

See `tutorials/auth_tutorial.Rmd` for full details. In short:

```{r auth, eval=FALSE}
api_key <- get_key()
hdr     <- auth_headers(api_key)
```

---

## Step 2 — Fetch project systems and select a procedure

`get_project_systems()` returns the full schema for your project.
`list_systems()` prints a readable summary of available systems and procedures.
`get_procedure()` extracts the specific procedure you want to upload against.

```{r project-systems, eval=FALSE}
project_systems <- get_project_systems(hdr)

# Print available systems and procedures.
list_systems(project_systems)

# Select the procedure that matches your CSV data.
procedure <- get_procedure(
  project_systems,
  system_name    = "Plante Ivindo",
  procedure_name = "Arbre"
)
```

---

## Step 3 — Validate the CSV

Run this before uploading to catch missing columns, bad values, and
unrecognised labels early. This is also where every label value gets
checked against the label database - which can take a while on a large
CSV, so you only want to do it once.

```{r validate, eval=FALSE}
csv_path <- "tutorials/data/example_observation_data_wide.csv"

# validate_csv_against_procedure() takes a data frame, not a path, so read
# it yourself first - free to filter/mutate it before validating.
observation_data <- readr::read_csv(csv_path, show_col_types = FALSE)

validated <- validate_csv_against_procedure(
  procedure        = procedure,
  observation_data = observation_data,
  hdr              = hdr
)

table(validated$status)
validated[validated$status == "rejected", c("status", "message")]
```

---

## Step 4 — Upload the validated result

Pass `validated` straight into `upload_observations_from_csv()` - it reuses
everything already built and checked, so nothing is rebuilt and no label is
checked against the database a second time. This is the only way to
upload: there's no direct `csv_path`/`procedure` shortcut, since validating
first is mandatory.

```{r upload, eval=FALSE}
result <- upload_observations_from_csv(
  hdr       = hdr,
  validated = validated
)

result$result
```

---

## Full tutorial script

See `tutorials/upload_observations_from_csv.R` for a complete, runnable example.
